---
title: "PS0002 Lab 2 Notes"
author: "Caleb Tay (caleb.tay@ntu.edu.sg)"
date: "6 Feb 2025"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    df_print: tibble
    template: ../eisvogel.latex
toc: true
toc-own-page: true
numbersections: true
papersize: a4
geometry:
- top=25mm
- left=25mm
- right=25mm
- bottom=25mm
- headsep=4mm
- footskip=10mm
editor_options: 
  markdown: 
    wrap: 80
---

# The `msleep` dataset

This lab uses the `msleep` dataset, which you can download using the
`download.file()` function and link below. Make sure that the link is on a
single line, so that R can read the link properly. Alternatively, you may also
navigate to the link and save the file directly to your computer.

```{r eval=FALSE}
download.file("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv",
  destfile = "msleep_ggplot2.csv")
```

Once downloaded, you need to import the data into R, with a use of one of the
`read` functions. Which one? The lab sheet suggests `read.csv()`. How can you
conclude by yourself what to use? Again, examine the data.

In this case, the file provided has a `.csv` extension, short for
*comma-separated values*. Opening the file, you can also see that each row
corresponds to one observation, and each feature is separated by a comma. There
is also a header row, so set the `header` parameter accordingly.

In the previous lab, I also mentioned that `read.csv()` is the same as
`read.table()`. You can in fact use `read.table()`, just change the `sep`
parameter.

```{r}
msleep <- read.csv("msleep_ggplot2.csv", header=T)  # equivalent to below
msleep <- read.table("msleep_ggplot2.csv", header=T, sep=",")
```

## Dimension (`dim()`), summary (`summary()`), and head (`head()`) of `msleep`

The lab sheet asks us to use the following functions.

```{r}
dim(msleep)
summary(msleep)
head(msleep)
```

From the `dim()` output, we can see that there are 83 rows and 11 columns.

The output of `summary()` gives us some summary statistics of the data. You can
see that it lists all the feature names, and where possible, numerical
statistics on the feature. For example, `sleep_total` has a minimum of 1.9, a
median of 10.10, a mean of 10.43, and a max of 19.90. For features that are
names or characters (strings), `summary()` cannot provide a mean, median, etc.
as those are numerical statistics.

The `head()` function simply gives the first few rows of the passed data frame.
You can optionally pass it a second number to explicitly determine how many rows
from the top it should take. Again, you can see this information with the `?`
operator.

## Selecting features (`dplyr`'s `select()`)

Most of these exercises are fairly simple. If you face an error doing this task,
make sure you have imported `dplyr` with `library()`.

```{r results='hide', warning=FALSE, message=FALSE}
library('dplyr')
select(msleep, name, sleep_total)
select(msleep, -(name:conservation))  # remove the character type data
select(msleep, -(1:4))              # remove the first four columns
msleep %>% select(sleep_total: awake)
msleep %>% select(starts_with("sl"))
```

Note the use of the pipe operator `%>%` is valid even with only a single verb.

## Filtering observations (`dplyr`'s `filter()`)

Again, this is an exercise of specifying the criteria to filter by accurately.

```{r results='hide'}
filter(msleep, sleep_total > 16)
filter(msleep, sleep_total > 16, bodywt > 1)
msleep %>% filter(order %in% c("Perissodactyla", "Primates"))
```

## Pipe operator (`%>%`)

The pipe operator allows for chaining of functions. It passes the left object as
the first input of the right function. If you want to spread it across multiple
lines, the pipe operator should be at the end of the line.

```{r results='hide'}
msleep %>% select(name, sleep_total) %>%    # <- pipe operator must be here if line break
  head(6)

# Equivalently, on a single line, without the pipe operator.
# Harder to understand due to its nested nature.
head(select(msleep, name, sleep_total), 6)
```

## Sorting data (`dplyr`'s `arrange()`)

```{r}
msleep %>% select(order, sleep_total, brainwt) %>%
  arrange(sleep_total) %>% # desc(sleep_total) if required
  filter(sleep_total > 16)
```

One thing to note about this chaining of operations is that the order of
operations may matter for large datasets. In this case, it does not matter
significantly whether the dataset is sorted or filtered first. One erroneous
example, where sorting by an unselected feature causes an error:

```{r eval=FALSE}
# cannot sort by brainwt, since select(msleep, sleep_total)
# only contains sleep_total, and not brainwt.
msleep %>% select(sleep_total) %>%
  arrange(brainwt)
```

In general, if you face an unexpected error or output, **break down your data
pipeline and see the output after each step**. That should give you a good idea
of where your code went wrong.
