---
title: "PS0002 Tut 1 Notes"
author: "Caleb Tay (caleb.tay@ntu.edu.sg)"
date: "21 Jan 2026"
output:
  pdf_document:
    df_print: tibble
    template: ../eisvogel.latex
  html_document:
    toc: true
    toc_float: true
toc: true
toc-own-page: true
numbersections: true
papersize: a4
geometry:
- top=25mm
- left=25mm
- right=25mm
- bottom=25mm
- headsep=4mm
- footskip=10mm
---

# Introduction to R

In PS0001, you learnt Python and its syntax. In this module, we use another
language, R. Generally speaking, R was built for statistical analysis, with many
built-in functions. Similar functionality in Python must be imported from
another library.

Why use R over Python? Well, it's just a different language.

Many of the same concepts overlap - variables, functions, iteration, etc.
However, a slight difference is that unlike in PS0001, we will actually be much
more hands-on with the running of our code - we are interested in running a
function, and seeing its output, immediately.

# The `cat()` and `print()` functions

The `cat()` function (short for con`cat`enate) and the `print()` function can be
used to display information to the user. Both, however, work slightly
differently.

We will not frequently use the `cat()` or `print()` functions, because we
generally see the output immediately.

```{r results='hide'}
cat('abc')
print('xyz')
```

# Assignment operator (`<-` or `=`)

`<-` and `=` can both be used to assign values to variables in R. For
consistency I will generally use `<-`.

```{r}
x = 1
y <- 2
cat('x has the value', x, ', y has the value', y)
```

# Vectors (`c()`)

Vectors are created using the `c()` function. When a range of consecutive
numbers is desired, we can also use `start:end` to get a vector of all numbers
between `start` and `end`, inclusive of both ends.

```{r}
v1 <- c(1, 3, 5, 7, 9, 11)
v2 <- 1:6*2   # create the vector (1, 2, 3, 4, 5, 6) and multiply each element by 2.
cat('v1 has the value', v1, ', v2 has the value', v2)
```

You can see the length of a vector using `length()`:

```{r}
cat('v1 has length', length(v1))
```

# Data frames (`data.frame`)

Data frames are what we will be handling throughout the rest of the course. They
are a collection of data, organised in **rows** and **columns**. They can be
created with the `data.frame()` function. In this case, there are 3 sets of
data.

```{r}
df <- data.frame(c(26, 24), c(175, 172), c("M", "F"))
df
```

The rows are individual observations, the columns are the data of each
observation. By default, the columns are not assigned any meaningful names, and
the rows are assigned numerical labels. We can see the column names and assign
new names using the `names()` (or `colnames()` function.

```{r}
names(df) # equivalently, colnames(df)
names(df) <- c('age', 'height', 'gender')
df
```

The column names could also be given directly in the `data.frame()` function:

```{r}
df1 <- data.frame(age = c(26, 24), height = c(175, 172), gender = c("M", "F"))
df1
```

We can also change the row names if desired using `rownames()`.

```{r}
rownames(df)
rownames(df) <- c('caleb', 'taylor')
df
```

## Dimension of a data frame (or matrix) (`dim()`)

`dim()` tells us the dimension of an object. A vector has no dimension, only
length. In this case, `dim()` tells us there are 2 rows and 3 columns.

```{r}
dim(df)
```

Now, to describe this data frame, I will say that this data frame has 2
observations (rows), and 3 data points (columns) for each observation. The
interpretation of this data frame with these row and column names should come
naturally based on the column and row names.

In this course, in general, we will not be creating data frames directly.
Instead, most data will be imported using a `read` function, which we will see
in the lab, or imported from another library, which we will see in a few weeks.

## Getting help (`?` and `help()`)

Many functions in R have documentation on them, i.e. what the function does, how
the function is used, the parameters it accepts. If you ever forget what a
function does or how to use it, in the R console you can use the `?` operator or
the `help()` function.

For example, typing `?names` or `help(names)` *without the parentheses* will
bring up the documentation page for the `names()` function.
You can do this with nearly anything, e.g `c()`, `dim()`, `data.frame()`

# Indexing vectors, matrices, and data frames

To get a single value in a vector, matrix, or data frame, we can use indexing
with square brackets `[]`. Indexing in R is extremely similar to indexing in
Python, but unlike Python, which is 0-indexed, R is 1-indexed, so the first
entry in a list always has index 1. Vectors in R are analogous to lists in
Python (that do not contain other lists), and require only one indexing.

```{r}
x <- 1:5*10
x
x[2]            # Get the second element
```

For matrices and data frames, as they have two dimensions (rows and columns), we
have the option to get rows, columns, or individual cells. The square bracket
notation is the same, but we now have the option to index along the rows or
columns or both. Take note of the below examples, noting the presence of the
commas and blank entries.

```{r}
df['age']     # with no comma, selects columns, return as *data frame*.
              # or with comma, first is row, second is column.
df[, 'age']   # selects all rows, and only column 'age', return as *vector*.
df['caleb',]  # selects only row 'caleb', and all columns, return as *data frame*.
```

Take note of the difference with and without the comma. Without the comma, it is
returned as a **data frame**. With the comma, it is returned as a **vector**.
This distinction will not be useful for a while, but some functions accept
**only vectors** or **only data frames**.

In general, the indexing is done as `thing_to_index[rows, columns]`. Named
columns in data frames can also be accessed using a `$` operator.

```{r}
df$age         # equivalent to df2[, 'age'], return as vector.
```

To select multiple rows or multiple columns, you can pass in vectors as indexes.
Selections can also be made using numerical indices. You can see that the below
have the exact same output as using the named indices above.

```{r}
df
df[1:2]        # get the first to second columns
df[1, ]        # get the first row, and all columns.
df[, c(1, 3)]  # get all rows, and the first and third columns
```

Indexing for vectors and matrices is extremely similar. As an exercise, index
the following values from the below matrix with a **single indexing operation**,
using the `c()` function if necessary.

```{r}
m1 <- matrix(1:20, nrow=4, byrow=T)
m1

# Index the following from the matrix m1:
# The entry 13
# The entries 1, 2, 3, 4, and 5.
# The entries 4, 9, 14, and 19.
# The entries 2, 7.
# The entries 1, 3, 16, and 18.

# For example, to index the entry 12
# which is the 3rd row, 2nd column.
m1[3, 2]

m1[3,3]
m1[1,]
m1[,4]
m1[c(1,2),2]
m1[c(1,4),c(1,3)]
```
