---
title: "PS0002 Tut 1 Notes"
author: "Caleb Tay (caleb.tay@ntu.edu.sg)"
date: "26 Dec 2025"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    df_print: tibble
    template: ../eisvogel.latex
toc: true
toc-own-page: true
numbersections: true
papersize: a4
geometry:
- top=25mm
- left=25mm
- right=25mm
- bottom=25mm
- headsep=4mm
- footskip=10mm
---

# Introduction to R

In PS0001, you learnt Python and its syntax.
In this module, we use another language, R.
Generally speaking, R was built for statistical analysis, with many built-in functions.
Similar functionality in Python must be imported from another library.

Why use R over Python?
Well, it's just a different language.

Many of the same concepts overlap - variables, functions, iteration, etc.
However, a slight difference is that unlike in PS0001, we will actually be much more hands-on with the running of our code - we are interested in running a function, and seeing its output, immediately.

# The `cat()` and `print()` functions

The `cat()` function (short for con`cat`enate) and the `print()` function can be used to display information to the user.
Both, however, work slightly differently.

We will not frequently use the `cat()` or `print()` functions, because we generally see the output immediately.

```r
cat('abc')
print('xyz')
```

# Assignment operator (`<-` or `=`)

`<-` and `=` can both be used to assign values to variables in R. For consistency I will just use `<-`.

```{r}
x = 1
y <- 2
cat('x has the value', x, ', y has the value', y)
```

# Vectors (`c()`)
Vectors are created using the `c()` function.
When a range of consecutive numbers is desired, we can also use `start:end` to get a vector of all numbers
between `start` and `end`, inclusive of both ends.

```{r}
v1 <- c(1, 3, 5, 7, 9, 11)
v2 <- 1:6*2   # create the vector (1, 2, 3, 4, 5, 6) and multiply each element by 2.
cat('v1 has the value', v1, ', v2 has the value', v2)
```

You can see the length of a vector using `length()`:

```{r}
cat('v1 has length', length(v1))
```

# Matrices
Matrices in R are similar to matrices in math.
They can be created in several ways.

## `dim()`
`dim()` tells us the dimension of an object.
A vector has no dimension, only length.
We can assign a dimension to a vector to make it a matrix.

```{r}
dim(v1)
dim(v1) <- c(3, 2)
v1      # now a 3x2 matrix
```

## `matrix()`
We can create matrices directly with `matrix()`.
Specify the parameter(s) `ncol`, `nrow`, and/or `byrow` to determine how the matrix should be created.

```{r}
matrix(1:10, nrow = 2)
matrix(1:10, ncol = 2)
matrix(1:10, ncol = 5, byrow=T)
```

## `rbind()` and `cbind()`
Matrices can also be made using `rbind()` and `cbind()` on vectors or other matrices.

```{r}
a <- matrix(1:10, nrow = 2)
b <- matrix(11:20, nrow = 2)
cbind(a, b)
rbind(a, b)

rbind(1:5, 6:10)
```

# Data frames (`data.frame`)
Data frames are just matrices with column names.
They can be created with the `data.frame()` function on any matrix or vector.

```{r}
df1 <- data.frame(1:5)
df1
a
df2 <- data.frame(a)
df2
```

The rows are individual observations, the columns are the data of each observation.
By default, the columns are assigned the names X1, X2, ..., Xn, for each column,
and the rows are assigned numerical labels.
We can see the column names, and assign new names using the `names()` function.

```{r}
names(df2)
names(df2) <- c('age', 'height', 'weight', 'grade', 'money')
df2
```

We can also change the row names if desired using `row.names()`.

```{r}
row.names(df2)
row.names(df2) <- c('caleb', 'you')
df2
```

Now, to describe this data frame,
I will say that this data frame has 2 observations (rows), and 5 data points (columns) for each observation.
The interpretation of this data frame with the row and column names should come naturally.

In this course, in general, we will not be creating data frames or matrices directly.
Instead, most data will be read using a `read` function, which we will see in the lab, or imported from another library, which we will see in a few weeks.

# Indexing vectors, matrices, and data frames
To get a single value in a vector, matrix, or data frame, we can use indexing with square brackets `[]`.
Indexing in R is extremely similar to indexing in Python,
but unlike Python, which is 0-indexed, R is 1-indexed, so the first entry in a list always has index 1.
Vectors in R are analogous to lists in Python (that do not contain other lists),
and require only one indexing.

```{r}
x <- 1:5*10
x
x[2]            # Get the second element
```

For matrices and data frames, as they have two dimensions (rows and columns), we have the option to get rows, columns, or individual cells.
The square bracket notation is the same, but we now have the option to index along the rows or columns or both.
Take note of the below examples, noting the presence of the commas and blank entries.

```{r}
df2['age']      # with no comma, selects columns, return as data frame.
                # or with comma, first is row, second is column.
df2[, 'age']    # selects all rows, and only column 'age', return as vector.
df2['caleb',]   # selects only row 'caleb', and all columns, return as data frame.
```

In general, the indexing is done as `thing_to_index[rows, columns]`.
Columns in data frames can also be accessed using a `$` operator.

```{r}
df2$age         # equivalent to df2[, 'age'], return as vector.
```

To select multiple rows or multiple columns, you can pass in vectors as indexes.
Selections can also be made using numerical indices.
You can see that the below have the exact same output as using the named indices above.

```{r}
df2[1:3]
df2[, c(2, 5)]
df2[1, ]
```

Indexing for vectors and matrices is extremely similar.
As an exercise, index the following values from the below matrix, using the `c()` function if necessary.

```{r}
m1 <- matrix(1:20, nrow=4, byrow=T)
m1

# Index the following from the matrix m1:
# The entry 13
# The entries 1, 2, 3, 4, and 5.
# The entries 4, 9, 14, and 19.
# The entries 2, 7.
# The entries 1, 3, 16, and 18.

# For example, to index the entry 12
# which is the 3rd row, 2nd column.
m1[3, 2]
```
